Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-04-27 drafts/hybrid_zone.tex
YB made minor changes to abstract and intro. Make sure to recompile [only pushed .tex]

\abstract{
Hybrid zones formed between recently diverged populations offer an opportunity 
to study the mechanisms underlying reproductive isolation and the process of speciation.
Here, we use a combination of analytical theory and explicit forward simulations
to describe how selection against hybrid genotypes impacts patterns of introgression
across genomic and geographic space. 
By describing how lineages move across the hybrid zone, in a model without coalescence, 
we add to modern understanding of how clines form and how parental haplotypes are broken up during introgression.
Working with lineages makes it easy to see that
clines form in about $1/s$ generations, where $s$ is the strength of selection against hybrids,
and linked clines persist over a genomic scale of $1/T$, where $T$ is the age, in generations, of the hybrid zone. 
Locally disadvantageous alleles tend to exist as small families,
whose lineages trace back to the side from which they originated at speed $\sqrt{s}$ dispersal distances per generation.
The lengths of continuous tracts of ancestry provide an additional source of information:
blocks of ancestry surrounding single-locus incompatibilities can be substantially longer than the genome-wide average block length at the same spatial location,
an observation that might be used to characterize the age of hybrid zones
and identify candidate targets of selection.
%\alisa{ascertainment bias. Tract lengths provide additional source of information, as they provide unbiased information even when used ascertained markers.}

@@

The process of speciation commonly involves populations diverging with relatively little gene flow \citep{Coyne2004}.  
However, when formerly isolated populations come into contact before reproductive isolation is complete, some gene flow is possible.
Interbreeding and migration between such populations  
creates a gradient of alleles derived from the two populations across geographic space,
centered on their point of contact \citep[reviewed by][]{Barton1985}. 
If the populations are sufficiently diverged, 
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
forming chromosomal {junctions} in ancestry \citep{Fisher1954, Chapman2002, baird2003distribution}. 
These patterns of correlated ancestry in admixed populations have been previously used to infer histories of hybridization and admixture 
\citep[e.g.,][]{Gravel2012,Hellenthal2014,Harris2013,sedghifar2015spatial}.

@@

Our work here builds on the large body of theory describing tension zones 
and aims to describe \emph{transient} dynamics with a focus on \emph{haplotype} patterns.
% which may carry substantial information about the formation of recent zones.
To do this, we take the reverse-time perspective,
understanding the temporal dynamics of the system by describing how lineages tracing back from the hybrid zone
moveme across space, conditioning on the frequencies of the selected allele and following them as they recombine between selected backgrounds
\citep[similar to][]{hudson1988coalescent}.
Using a combination of theory and simulated hybrid zones, 
and extending the framework of \citet{sedghifar2015spatial},
we describe genome-wide patterns of coancestry 
as they relate to hybrid zone age, genetic distance from selected locus and geographic distance from hybrid zone center.

@@

We neglect genetic drift in our analytical results \revpoint{2}{9}
because of the well-known problems with spatial coalescence \citep{felsenstein1975pain,barton2002continuous}.
This should provide a good model for stochastic motion of a single lineage (our main focus),
but does not model correlations between lineages or non-Markovian dynamics of the lineages.
This approach is common to much theoretical work,
that studies the reaction-diffusion equations that govern the deterministic, high-density limit \citep[as in][]{Nagylaki1975},
or the discrete analogues \citep{hanson1966effects}. 
There is also substantial work on how clines are affected by genetic drift \citep{Slatkin1975,felsenstein1975genetic,Nagylaki1978,durrett2007width,barton2008effect,Polechova2011}.
Our simulation results (below) suggest that although genetic drift strongly affects local noisiness of the system,
it does not substantially influence our conclusions at the moderate population densities we examine,
perhaps because we study relatively short time-scales.
At much lower population densities, however, local extinction and recolonization may have a stronger effect. \linelabel{ll:much_lower_densities}

>>>>

\abstract{
Hybrid zones formed between recently diverged populations offer an opportunity 
to study the mechanisms underlying reproductive isolation and the process of speciation.
Here, we use a combination of analytical theory and explicit forward simulations
to describe how selection against hybrid genotypes impacts patterns of introgression
across genomic and geographic space. 
By describing how lineages move across the hybrid zone, in a model without coalescence, 
we add to modern understanding of how clines form and how parental haplotypes are broken up during introgression.
Working with lineages makes it easy to see that
clines form in about $1/s$ generations, where $s$ is the strength of selection against hybrids,
and linked clines persist over a genomic scale of $1/T$, where $T$ is the age, in generations, of the hybrid zone. 
Locally disadvantageous alleles tend to exist as small families,
whose lineages trace back to the side from which they originated at speed $\sqrt{s}$ dispersal distances per generation.
The lengths of continuous tracts of ancestry provide an additional source of information:
blocks of ancestry surrounding incompatibilities can be substantially longer than the genome-wide average block length at the same spatial location, %single-locus incompatibilities can be substantially longer than the genome-wide average block length at the same spatial location, % YB Comment. I think single-locus is a red-herring as we hope this will work for singe loci involve in multi-locus incompatibilities
an observation that might be used to % characterize the age of hybrid zones and %YB Comment - age seems like our last paper?
identify candidate targets of selection.
%\alisa{ascertainment bias. Tract lengths provide additional source of information, as they provide unbiased information even when used ascertained markers.}

@@

The process of speciation commonly involves populations diverging with relatively little gene flow \citep{Coyne2004}.  
However, when formerly isolated populations come into contact before reproductive isolation is complete, some gene flow is possible.
Interbreeding and migration between such populations  
creates a gradient of alleles derived from the two populations across geographic space,
centered on their point of contact \citep[reviewed by][]{Barton1985}. 
If the populations are sufficiently diverged, 
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
forming chromosomal {junctions} in ancestry \citep{Fisher1954, Chapman2002, baird2003distribution}. 
These patterns of correlated ancestry in admixed populations have been previously used to infer histories of hybridization and admixture 
\citep[e.g.,][]{Gravel2012,Hellenthal2014,Harris2013,sedghifar2015spatial}. 
Here we show how natural selection against hybrid incompatibilities shapes patterns of coancestry around such loci. %YB Comment. I really want people to know what we are doing by the end of the 1st paragraph

@@

Our work here builds on the large body of theory describing allele frequency clines in tension zones % YB added ' allele frequency clines'
and aims to describe \emph{transient} dynamics with a focus on \emph{haplotype} patterns.
% which may carry substantial information about the formation of recent zones.
To do this, we take the reverse-time perspective,
understanding the temporal dynamics of the system by describing how lineages tracing back from the hybrid zone
moveme across space, conditioning on the frequencies of the selected allele and following them as they recombine between selected backgrounds
\citep[similar to][]{hudson1988coalescent}.
Using a combination of theory and simulated hybrid zones, 
and extending the framework of \citet{sedghifar2015spatial},
we describe genome-wide patterns of coancestry 
as they relate to hybrid zone age, genetic distance from selected locus and geographic distance from hybrid zone center.

@@

To address this idea we develop analytical theory and a set of stochastic simulations. % YB Comment. i thought a paragraph starting with neglect was a downer
Our analytical theory provides a quantitative description of the problem, but in so doing, we neglect genetic drift  \revpoint{2}{9}
because of the well-known problems with spatial coalescence \citep{felsenstein1975pain,barton2002continuous}.
This should provide a good model for stochastic motion of a single lineage (our main focus),
but does not model correlations between lineages or non-Markovian dynamics of the lineages.
This approach is common to much theoretical work,
that studies the reaction-diffusion equations that govern the deterministic, high-density limit \citep[as in][]{Nagylaki1975},
or the discrete analogues \citep{hanson1966effects}. 
There is also substantial work on how clines are affected by genetic drift \citep{Slatkin1975,felsenstein1975genetic,Nagylaki1978,durrett2007width,barton2008effect,Polechova2011}.
Our simulation results  suggest that although genetic drift strongly affects local noisiness of the system,
it does not substantially influence our conclusions at the moderate population densities we examine,
perhaps because we study relatively short time-scales.
At much lower population densities, however, local extinction and recolonization may have a stronger effect. \linelabel{ll:much_lower_densities}
